---
title: 'Drugs and the Police: San Francisco and Seattle'
author: "Stuart Barnum"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

Examining the Seattle and San Francisco crime data for the summer of 2014, as reported by the cities, I noticed that San Francisco, despite having a population less than 30 percent greater than the population of Seattle, had more than twice as many recorded cases of drug crime. Does this reflect a greater rate of use of certain drugs in San Francisco than in Seattle, or, rather, perhaps a difference in the manner of enforcement of drug laws or the manner in which reported infractions are recorded? Digging somewhat deeper---examining the data from 2013 through 2017 and on the use of certain popular drugs---I uncovered some striking trends and obtained some possible insight into these questions.

Thus examining the Seattle and San Francisco crime data from 2013 through 2017, one of the first things I noticed was a dramatic increase in the number of recorded incidents of crime. For example, in 2017, the number of recorded cases of drug crime in Seattle were, in contrast to the numbers for 2014, greater than the number recorded cases of drug crime in San Francisco. As the following plot of incidents of crime within some selected categories suggests, the fractional increase in recorded incidents of crime in Seattle is roughly the same across the various categories of crime.

The San Francisco dataset was downloaded on 2/7/2018, at:
`https://data.sfgov.org/Public-Safety/Police-Department-Incidents-Current-Year-2017-/9v2m-8wqu`

Seattle dataset was downloaded on 2/7/2018, at: `https://data.seattle.gov/Public-Safety/Seattle-Police-Department-Police-Report-Incident/7ais-f98f/data`

```{r,}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, cache = TRUE}
SF_complete <- read_csv("SF_Complete.csv")

#With guess max large enough, columns that contain integers that exceed the 
#32-bit maximum are read as character vectors.
seattle_complete <- read_csv("seattle_complete.csv", guess_max = 100000)

```

```{r}
library(tidyverse)
library(lubridate)

SF_complete <- SF_complete %>%
  mutate(Year = as.integer(str_sub(Date, -4, -1))) %>%
  mutate(Category = fct_recode(Category,
        "Drugs" = "DRUG/NARCOTIC"))

seattle_complete <- seattle_complete %>%
  rename(Category = `Summarized Offense Description`) %>%
  mutate(Category = fct_recode(Category,
        "Prostitution" = "PROSTITUTION",
        "Drugs" = "NARCOTICS",
        "Weapons" = "WEAPON",
        "Liquor\nlaws" = "LIQUOR VIOLATION",
        "Assault" = "ASSAULT",
        "Homicide" = "HOMICIDE",
        "Robbery" = "ROBBERY",
        "Vehicle\ntheft" = "VEHICLE THEFT",
        "Theft\nfrom\nvehicle" = "CAR PROWL"))

seattle_complete <- seattle_complete %>%
  mutate(R_date = floor_date(mdy_hms(`Occurred Date or Date Range Start`,
                                     tz = "PST8PDT"), "day"))

SF_complete <- SF_complete %>%
  mutate(R_date = mdy(Date, tz = "PST8PDT"))

seattle_complete %>%
  filter(Category %in% 
        c("Prostitution",
        "Drugs",
        "Weapons",
        "Assault",
        "Homicide",
        "Robbery",
        "Vehicle\ntheft",
        "Theft\nfrom\nvehicle")) %>%
  filter(Year %in% 2013:2017) %>%
  ggplot(aes(x = Category, fill = as.factor(Year))) +
  geom_bar(position = "dodge") +
  ggtitle("Recorded Cases of Selected Categories of Crime: Seattle") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Year")

```

San Francisco has not seen such a sharp increase in the number of recorded incidents. Furthermore, there is no evidence of the massive crime wave in Seattle that this data would suggest if the increase were taken to reflect the actual increase in crime. For example, note that the graph shows an increase in recorded instances of drug crime, from 2016 to 2017, of more than 100 percent. However, although there are credible reports of recent increases in use of some drugs, particularly heroin (see http://adai.uw.edu/pubs/pdf/2016drugusetrends.pdf), none of the reports indicates an increase  approaching anything close to 100 percent.

Looking somewhat deeper at the recorded cases of illegal drug use in the two cities, consider the differences between crimes that would be considered trafficking (sales, transportation, and, in my schema, production) and those that would be considered possession.


```{r, fig.height = 5, fig.width = 10}
library(plotly)

seattle_drugs_13_17 <- seattle_complete %>%
  filter(Category == "Drugs" & Year %in% c(2013:2017))

SF_drugs_13_17 <- SF_complete %>%
  mutate(`Cited or Arrested` = (str_detect(Resolution, "ARREST") | 
                                str_detect(Resolution, "CITED"))) %>%
  filter(Category == "Drugs" & Year %in% c(2013:2017))

SF_drugs_13_17 <- SF_drugs_13_17 %>% 
  mutate(`Drug Offense Type` = 
          ifelse(str_detect(Descript, "LAB APPARATUS"), "Sales, Transportation, or Production",
              ifelse(str_detect(Descript, "POSSESSION"), "Possession",
              ifelse(str_detect(Descript, "SALE"), "Sales, Transportation, or Production",
              ifelse(str_detect(Descript, "TRANSPORT"), "Sales, Transportation, or Production",
              ifelse(str_detect(Descript, "PLANTING/CULTIVATING MARIJUANA"), 
                     "Sales, Transportation, or Production",
              "Other"
              ))))))

seattle_drugs_13_17 <- seattle_drugs_13_17 %>%
  mutate(`Drug Offense Type` = 
          ifelse(str_detect(`Offense Type`, "POSSESS"), "Possession",
              ifelse(str_detect(`Offense Type`, "FOUND"), "Possession",
              ifelse(str_detect(`Offense Type`, "DISTRIBUTE"), 
                     "Sales, Transportation, or Production",
              ifelse(str_detect(`Offense Type`, "SMUGGLE"), 
                     "Sales, Transportation, or Production",
              ifelse(str_detect(`Offense Type`, "PRODUCE"), 
                     "Sales, Transportation, or Production",
              ifelse(str_detect(`Offense Type`, "SELL"), 
                     "Sales, Transportation, or Production",
              ifelse(str_detect(`Offense Type`, "TRAFFIC"), 
                     "Sales, Transportation, or Production",
              "Other"))))))))
  

joined_counts <- (seattle_drugs_13_17 %>% count(Year, `Drug Offense Type`)) %>%
  left_join((SF_drugs_13_17 %>% count(Year, `Drug Offense Type`)), by = c("Year", "Drug Offense Type")) %>%
  gather(n.x, n.y, key = City, value = Incidents) %>%
  mutate(City = fct_recode(City,
                           Seattle = "n.x",
                           `San Francisco` = "n.y"))

#reformat some categories that will appear in the plot
joined_counts <- joined_counts %>% 
  mutate(`Drug Offense Type` = as.factor(`Drug Offense Type`)) %>%
  mutate(`Drug Offense Type` = 
           fct_recode(`Drug Offense Type`,
                      " Sales, Transportation,\nor Production" = "Sales, Transportation, or Production",
                      " Possession" = "Possession"))

plot <- ggplot(joined_counts %>% filter(`Drug Offense Type` != "Other")) +
  geom_col(aes(x = Year, y = Incidents, fill = City, color = `Drug Offense Type`),
                        position = "dodge") +
  theme(legend.title = element_blank()) +
  labs(x = "year", y = "count",
       title = "Drug-related Police Incidents: possession versus trafficking")

ggplotly(plot) %>% layout(margin = list(b = 50, l = 60, r = 10, t = 80))
```

We see that, unlike the Seattle data, the San Francisco data show a steady decrease in recorded drug-crime incidents from one year to the next. The recorded incidents for Seattle show similar fractional decreases from 2013 to 2014, but then, as with the data related to other categories of crime, begin to increase significantly after that. Let's look at the data for particular drugs, and consider it together with the totals for all of the recorded instances of crime.


```{r, fig.height = 5, fig.width = 10}
seattle_drugs_13_17 <- seattle_drugs_13_17 %>%
  mutate(`Drug Type` = ifelse(str_detect(`Offense Type`, "MARIJU"), "Marijuana",    #yes
                            ifelse(str_detect(`Offense Type`, "METH"), "Methamphetamine",    #yes
                            ifelse(str_detect(`Offense Type`, "COCAINE"), "Cocaine",  #yes 
                            ifelse(str_detect(`Offense Type`, "HEROIN"), "Heroin",  #yes
                            ifelse(str_detect(`Offense Type`, "PRESCRIPTION"), "Prescription", #?
                            ifelse(str_detect(`Offense Type`, "PILL/TABLET"), "Pill/Tablet",  #?
                            ifelse(str_detect(`Offense Type`, "HALLUCINOGEN"), "Hallucinogen",
                            ifelse(str_detect(`Offense Type`, "SYNTHETIC"), "Synthetic",  #?
                            ifelse(str_detect(`Offense Type`, "AMPHETAMINE"), "Amphetamine", #yes
                            ifelse(str_detect(`Offense Type`, "OPIUM"), "Opium", #yes
                            ifelse(str_detect(`Offense Type`, "PARAPHENALIA"), "Paraphernalia", #yes
                            "Other")
                            )))))))))))
  
SF_drugs_13_17 <- SF_drugs_13_17 %>%
  mutate(`Drug Type` = ifelse(str_detect(Descript, "MARIJUANA"), "Marijuana",   #yes
                            ifelse(str_detect(Descript, "COCAINE"), "Cocaine",
                            ifelse(str_detect(Descript, "METH-AMPHETAMINE"), "Methamphetamine",
                            ifelse(str_detect(Descript, "BARBITUATES"), "Barbituates",
                            ifelse(str_detect(Descript, "CONTROLLED SUBSTANCE"), 
                                              "Controlled Substance",
                            ifelse(str_detect(Descript, "HALLUCINOGENIC"), "Hallucinogenic",
                            ifelse(str_detect(Descript, "AMPHETAMINE"), "Amphetamine", 
                            ifelse(str_detect(Descript, "METHADONE"), "Methadone",
                            ifelse(str_detect(Descript, "PARAPHERNALIA"), "Paraphernalia",
                            ifelse(str_detect(Descript, "OPIATES"), "Opiates",
                            ifelse(str_detect(Descript, "OPIUM"), "Opium",
                            ifelse(str_detect(Descript, "HEROIN"), "Heroin",
                            "Other")
                            ))))))))))))


```



```{r, fig.height = 9, fig.width = 12}

library(gridExtra)

seattle_selected_drugs <- seattle_drugs_13_17 %>%
  filter(`Drug Type` %in% c("Marijuana", "Cocaine", "Methamphetamine", "Heroin"))

SF_selected_drugs <- SF_drugs_13_17 %>%
  filter(`Drug Type` %in% c("Marijuana", "Cocaine", "Methamphetamine", "Heroin"))

number_of_bins <- 78

plt_seattle_selected_drugs <- ggplot(seattle_selected_drugs) + 
  geom_freqpoly(aes(x = R_date, color = `Drug Type`), bins = number_of_bins) +
  scale_x_datetime(date_breaks = "4 months", 
                   limits = c(as.POSIXct("2012/12/31", tz = "PST8PDT"), 
                              as.POSIXct("2018/01/01", tz = "PST8PDT"))) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1)) +
  labs(y = "number per fortnight", x = "",
       title = "Seattle Police Incidents Related to 4 Popular Drugs",
       color = "Drug: ") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 14),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12)) +
  guides(size = guide_legend(order = 2))

plt_SF_selected_drugs <- ggplot(SF_selected_drugs) +
  geom_freqpoly(aes(x = R_date, color = `Drug Type`), bins = number_of_bins) +
  scale_x_datetime(date_breaks = "4 months", 
                   limits = c(as.POSIXct("2012/12/31", tz = "PST8PDT"), 
                              as.POSIXct("2018/01/01", tz = "PST8PDT"))) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1)) +
  labs(y = "number per fortnight", x = "",
       title = "San Francisco Police Incidents Related to 4 Popular Drugs") +
  theme(legend.position = "none",
        plot.title = element_text(size = 14),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 10))

plt_seattle_all_13_17 <- seattle_complete %>% 
  filter(Year %in% 2013:2017) %>%
  ggplot(aes(x = R_date)) + geom_freqpoly(bins = number_of_bins) +
  scale_x_datetime(date_breaks = "4 months", 
                   limits = c(as.POSIXct("2012/12/31", tz = "PST8PDT"), 
                              as.POSIXct("2018/01/01", tz = "PST8PDT"))) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1)) +
  labs(y = "number per fortnight", x = "",
       title = "Seattle Police Incidents: 2013 through 2017") +
  theme(plot.title = element_text(size = 14),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 10))

plt_SF_all_13_17 <- SF_complete %>%
  filter(Year %in% 2013:2017) %>%
  ggplot(aes(x = R_date)) + geom_freqpoly(bins = number_of_bins) +
  scale_x_datetime(date_breaks = "4 months", 
                   limits = c(as.POSIXct("2012/12/31", tz = "PST8PDT"), 
                              as.POSIXct("2018/01/01", tz = "PST8PDT"))) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1)) +
  labs(y = "number per fortnight", x = "",
       title = "San Francisco Police Incidents: 2013 through 2017") +
  theme(plot.title = element_text(size = 14),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 10))

grid.arrange(plt_seattle_all_13_17, plt_SF_all_13_17, 
             plt_SF_selected_drugs, plt_seattle_selected_drugs,  
             layout_matrix = rbind(c(1, 2),
                                   c(4, 3),
                                   c(4, NA)),
             heights = c(4, 6, .7))
             
```

This final graphic shows a sharp increase in recorded cases of crime in Seattle during 2016, of roughly 200 percent in a period of four months in the late summer and fall. The sharpness of this increase, which would not have corresponded to a similarly sharp increase in crime, likely corresponds to a change either in policing patterns or in systems for recording crime. We must thus, again, treat any apparent increase in drug-related crime in Seattle during this period with great care. However, it is striking that, during this four-month period, the number of recorded cases of marijuana crime did not increase significantly. If we assume, plausibly, that marijuana crime in Seattle did not decrease substantially in this period, we may infer that the emphasis by the Seattle police on the enforcement of marijuana laws actually decreased during this period.

The trend in San Francisco is very different. Here we see a rather steady rate of recorded cases of crime involving heroin, whereas the crime related to the other three popular drugs---cocaine, marijuana, and methamphetamine---show steady rates of decrease.



